Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2018-10-04 versions/response.v1.txt

> Reviewer: 1

> Comments to the Author
> The authors show that when ancestral states are considered as
> parameters, there are zones where topologies and edge-lengths are
> inconsistently estimated. As is evidenced by any ML package, it is
> widely accepted that ancestral states should not be parameters; the
> large number of parameters suggests the approach will be problematic.
> Still, to the best of my knowledge, it has not been shown that
> treating ancestral states as parameters actually leads to poor
> performance, or how poor. Moreover, the authors point to Sagulenko et
> al (2018) - now published in Virus Evolution - as an example
> implementation. I found the paper interesting and I believe a more
> succinct presentation will be of interest.

> A main concern with the paper is that the shaded regions of
> topological inconsistency or poor estimation in Figs 2-4, are only
> subsets of the actual regions where topological inconsistency or poor
> estimation occur; and yet determining the actual regions seems
> feasible. The contrast between Fig 5 and 4 illustrates how far off the
> subsets are in giving the actual regions where estimation is
> problematic. The white region in Fig 5 is the actual region where hat
> w=1 whereas the shaded region of Fig 3 is the subset of the actual
> region that the authors were able to determine through analytical
> arguments. For most readers, out of Fig 3 and Fig 5, only Fig 5 is of
> interest.

Thank you for the opportunity to improve on our analytically-derived regions of inconsistency.
Indeed, obtaining these regions in closed form is feasible, though it involves comparing a large number of two-variables functions to one another.
Fig. S2 shows the regions where a particular likelihood equation will be maximal, and allows us to compute---in closed form---the value of the branch parameter in Fig. 2.

@@

> Getting exact regions is what I think the authors should do. The
> approach used to obtain Fig 5 is generally applicable. Explicit,
> relatively simple, expressions are given in the appendix for long-run
> lnls, l_{tau_1,t*}(tau_2,w,x,y). For any given (x*,y*) on a fine grid
> in Figs 2-4, l_{tau_1,t*}(tau_j,w,x,y), can be maximized using
> numerical algorithms over the three parameters w,x,y. Doing so for
> tau_j=tau_1, tau_2 would determine the actual regions of problematic
> estimation corresponding to Figs 2-4. The only potential difficulty
> that I can see is that numerical algorithms might not converge. This
> can either be avoided by considering a cubic grid of starting (w,x,y),
> or through something like the following. For any given set of
> ancestral character state assignments for the distinct set of 8
> patterns 1111, 1110,... (0000 can treated as the same as 1111), the
> long-run lnl is of the form

> l(x,y,w)=a log(1-x) + b log(1+x) + c log(1-y) + d log(1+y) + e log(1-w) + f log(1+w)

> where a,..f>=0 are constants determined by (x*,y*) and what ancestral
> states have been assigned. The function l(x,y,w) has an explicit
> simple maximizer. Thus the overall maximizer is determined as the
> largest maximized l(x,y,w) over all 4^8 possible ancestral state
> assignments: a finite number of comparisons. The actual number of
> possible ancestral state assignments is quite a bit smaller than 4^8
> (for instance 11 is always the ancestral reconstruction for 1111) and
> the authors have a lot of familiarity with what the ancestral
> reconstructions are. I believe Fig 2 can be replaced with an exact (up
> to grid approximation) region of topological inconsistency and Figs
> 3-5 with heat maps of limiting (x,y,w) for tau_1.

Another thanks for this suggestion.
As noted, we can in fact get exact regions, and have done so.
Our line of argument starting on page 25, line 541 in the section "Properties of the joint objective function" allows us to pare down the 4^8 possible ancestral states to just 81, and we obtain exact, analytic results in a reasonable amount of time.

@@

> I know that it's not feasible to explicitly compute (analytically)
> these likelihoods for bigger trees. But one of the main motivations
> and applications for tools like TreeTime is surely that the number of
> sequences grows (not the number of sites, which is fixed by the viral
> sequence length in the case of viruses). What happens when the number
> of tips grows? Are there more opportunities for the kind of issue that
> gives the wrong value for w? Many influenza trees have many 0 (or near
> 0) internal branch lengths, and it would be interesting to understand
> the implications for the ancestral state reconstructions (which in
> turn are used to infer spatial spread, transmission etc).

%EM David-- does it seem like we could extend these results to bigger trees containing a 4 taxon tree?
%das: I wasn't able to come up with anything interesting for this.

> Some detailed comments and questions:
> – line 85 some cut/paste or sentence issue

We have added the definition of a "generating probability" to ease confusion in this sentence:
"As data becomes plentiful, we use the Hadamard transform (see (8) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."

@@

> line 364 define p(e)

Thanks for reminding us to define p(e), which is now described as "where $p(e)$ is the probability of a character change along edge $e$."

> line 366  I don't understand (8) but perhaps that's just background I
> don't have – a quick explanation or reference would be handy
> What's a 'generating probability' ? Is that the fidelity? Or a
> generator?

We have updated the text to explicitly define "generating probability."
"As data becomes plentiful, we use the Hadamard transform (see (8) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."

@@

> Comments to the Author
> Shaw and Matsen show that a joint maximum-likelihood inference of
> phylogenies and ancestral states is not consistent when branch lengths
> are long. We have recently proposed such a joint estimation for
> sequences that recently diverged and are hence similar (Sagulenko et
> al). The results presented here are anticipated in Sagulenko et al
> "For more divergent sequences, however, subleading states of internal
> nodes make a substantial contribution and the iterative optimization
> will underestimate the branch lengths."
> and
> "The iterative optimization procedures are not expected to be accurate
> for trees were many sites are saturated. In scenarios with extensive
> uncertainty of ancestral states and tree topology, convergence of the
> iterative steps cannot be guaranteed."

> In this light, it should be pointed out that joint estimation
> procedure was explicitly suggested as an approximation and hence never
> expected to be consistent in a strict sense. Furthermore, it was
> suggested as an approximation when branch lengths are short -- the
> opposite of the problematic regions identified here.
> But systematic analysis of the behavior of this approximation is
> useful and I believe the work is correct (but haven't checked the
> proofs). The work by Shaw and Matsen suggests the radius of
> convergence of this procedure is somewhere around branch length of
> order 10-20%, which is consistent with results reported by Sagulenko
> et al.

Thank you for your thoughts, and we have worked hard to not oversell our results.
The TreeTime paper does present joint optimization as an approximate approach, and described that it is for short branch lengths.
However, in our assessment the word "approximate" doesn't completely describe a situation in which there is significant bias.
The TreeTime paper says that its use is appropriate when "ancestral sequence inference has little ambiguity" but even in these cases the bias in branch lengths persists.

TODO we should make some changes to at least say that we've further clarified the implications of our results.
%das: I believe this is done
We clarify our aim and warnings from Sagulenko et al. (2017) in the Introduction:
"Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference."

@@

> At the same time, these results are not particularly surprising.
> Assigning ML ancestral states will always result in shorter branches.
> For short branches, this correction is linear in branch length, for
> longer branches higher order effects start to contribute. If the
> typical branch length around a bifurcating node is x<<1, a fraction ~x
> of the sites mutated along one of the three adjacent branches, while a
> fraction ~x^2 mutated on two branches and not the third. The latter
> case is ignored in this joint inference procedure. Not having to sum
> over these subleading states is what makes joint inference faster, but
> such truncation will first lead to inaccuracies and at longer
> distances to qualitatively different solutions (like Taylor expanding
> 1/(1-x) ).
>
> There are other obvious cases where joint inference will fail. For
> sequences with a heavily biased in nucleotide composition (say biased
> towards A). For sufficiently long terminal branches, all ancestral
> states will be estimated as all 'A' and subsequent branch length
> estimate will shrink all internal branches to zero. This happens for
> the very same reason that the most likely outcome of multiple flips of
> a biased coin is all heads -- a very atypical yet most likely outcome.
>
> Currently, the work largely shows that areas of inconsistencies exist.
> It would be useful if the authors could present Taylor expansions for
> the ML branch length estimates around x=y=w=1. This would more
> explicitly show the inaccuracies accumulate with increasing branch
> length and how the inaccuracies depend on asymmetries in the tree. To
> me, this would be much more useful than to know that joint inference
> will fail if branches are long.

This is an interesting suggestion but is different than our goal, which was to establish inconsistency.
Moreover, in the revised manuscript we see cases where assigning ML ancestral states results in longer branches, not shorter ones (see the discussion after the proof of Theorem 1, page 31, line 627).
We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree:
"This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\boldsymbol\xi=\emptyset$ in Table~\ref{tab:likelihoods})."

Asymmetries can drive these particular choices, but Fig. 2 in the revised manuscript shows that almost half of the parameter space results in an interior branch length being estimated as zero.
We do not think providing an argument about accumulating inaccuracies would elucidate why such starkly-defined regions of inconsistency exist.
For reasonable parameter settings, one could investigate the error accumulation given fixed ancestral states, but this was not our goal and will require another paper.

@@

Thank you for bringing this to our attention.
We have tried to use branch lengths in as many instances as we could in the discussions, using fidelities only in the computations as they allow for great simplifications.

>  - Fig 5: it is surprising that \hat{w} seems completely flat as a
>  function of x* until it jumps discontinuously to 0. Why not include
>  the analytical bound in Fig 5?

The discontinuity is a reflection of the change in ancestral states when the generating parameters change.
As the true generating parameters change, we become more likely to choose a different ancestral state as most likely, resulting in an entirely different objective function to be optimized.
We have added the following:
"There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal."

In the revised manuscript, all region bounds are analytical as we have determined a closed form solution for the joint inference problem in this case.

@@

>  - It would be nice if the abstract pointed out that these
>  inconsistencies arise when branch lengths are long and that the
>  approximate joint maximization procedure was suggested for precisely
>  the opposite limit. Similar at the beginning of the discussion, I
>  think it would be more appropriate if this would say smth along the
>  lines "Sagulenko et al suggested an approximation for short
>  branches... we show that this doesn't work for long branches..."

TODO we should do some change, but make it clear that it's consistently biased even for short branches.
%das: I believe this is done
We have added the line:
"Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference."

@@

It's also a good idea to cite additional papers about identifiability, and we have done so in the Introduction:
"An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood."
and Discussion sections:
"Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood."

%EM David, heh, can you point me to the place where the equal branch length result was done?
%EM Also, I think you said that he says that the varying branch length thing is an open question-- it would be great to quote that.
%das: update I added some of the below. Let me know what you think. And re: Allman/Chai, I just added them in with RoyChoudhury; is that ok? They are all just examples of conditions where phylogenetics is consistent it seems.

%das: below are some notes and quotes. I will work on paring these down and adding some to the text, maybe with a little more diplomacy.
%The equal branch length example occurs in Example 2, starting halfway down the second column of page 352 in Goldman (1990) and continuing until the top of the second column on page 354.
%Here he only talks about how parsimony is inconsistent because it may result in equally parsimonious trees (this is the last paragraph of the example section).
%It's not until the bottom of page 356 that he punts the inconsistency to Felsenstein and Hendy/Penny, and the bottom of page 357 where he wonders how small branch lengths have to be for parsimony to be consistent in general.

%das: @EM let me know if, given these notes, you think the discussion in the text is reasonable

### 2018-10-04 appendix.tex
pass thru
Quick look
pass through comments

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

Examples of generating parameters resulting in consistency of joint inference occur in the following.
Let $x^*$ and $y^*$ be such that the estimate for $\hat{x}_1$ is
\[
\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{123} + p_{13}\right) - 1 = (x^*)^2.
\]
There may be no $x^*$ and $y^*$ that yield this maximum (Fig.~\ref{fig:max-anc-state}), though it is a feasible solution to the general joint inference problem.
In this case, $\hat{x}_1 = x^*$ implies consistency only when $x^*=0$ or $x^*=1$.
Another case,
\[
\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{3} + p_{123} + p_{23} + p_{13}\right) - 1 = \frac{1}{2} + \frac{1}{2}(x^*)^2
\]
shows $\hat{x}_1 = x^*$ implies consistency only when $x^*=1$.

@@

\begin{figure}
    \begin{minipage}{\linewidth}
        \centering
        %file: plotting/_output/invfels/ancestral_state_conditions/non-bias/plot.png
        \includegraphics[width=.95\textwidth]{analytic-anc-state}
        \label{ }
    \end{minipage}
    \begin{minipage}{\linewidth}
        \centering
        Maximal ancestral state split definitions
        \begin{tabular}{l}
        \hline
        $\hat{\boldsymbol\xi} = \{\xi_{\emptyset},\xi_{1},\xi_{2},\xi_{3},\xi_{123},\xi_{12},\xi_{23},\xi_{13}\}$\\
        \hline
        $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$\\
        $\hat{\boldsymbol\xi}_2 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\emptyset\}$\\
        $\hat{\boldsymbol\xi}_3 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\{1\}\}$\\
        $\hat{\boldsymbol\xi}_4 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\{1,2\},\{1\}\}$\\
        $\hat{\boldsymbol\xi}_5 = \{\emptyset,\emptyset,\emptyset,\{1\},\{1\},\emptyset,\{1\},\{1\}\}$\\
        \hline
        \end{tabular}
    \end{minipage}
\caption{
Regions of maximal ancestral state splits on the InvFels tree $\tau_1$.
}
\label{fig:max-anc-state}
\end{figure}

@@

The regions in Fig.~\ref{fig:max-anc-state} are analytically-derived regions of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
As the region of degeneracy in Fig.~\ref{fig:max-anc-state} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
%EM "letting $\emptyset$ be the maximal ancestral state split" sounds a little funny-- either it is or it isn't maximal.
It is easy to see that when $\emptyset$ is the maximal ancestral state split, we have the same log-likelihood for $\tau_1$ and $\tau_2$.
Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the region in Fig.~\ref{fig:max-anc-state} where $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_1$.
We can see this results in the likelihood of both topologies being equal, i.e.,
\begin{align*}
\max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
\end{align*}

>>>>

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
%EM @das I think that ^ is all resolved, right?
%das: I think to be fair I should add that we don't actually show a real inconsistency in topology because we haven't shown this functional convergence for it, only for branch lengths with a fixed topology. I added a paragraph hedging a bit. Let me know what you think.
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

This Wald-type consistency does not extend readily to convergence of topology, i.e., we cannot use the above arguments to say that, for
\[
(\hat{\tau}_\nCols, \hat{t}_\nCols) = \arg\max_{\tau, t} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)
\]
then
\[
(\hat{\tau}_\nCols, \hat{t}_\nCols) \rightarrow (\hat{\tau}, \hat{t})
\]
as we did for $\hat{t}_\nCols$ with $\tau$ fixed due to the properties of tree space \citep{Yang1994}.
For this reason we do not claim joint inference is inconsistent in estimating topology, only that there exist parameters that yield degenerate maxima when performing joint optimization.

@@

Examples of generating parameters resulting in consistency of joint inference occur in the following.
Let $x^*$ and $y^*$ be such that the estimate for $\hat{x}_1$ is
\[
\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{123} + p_{13}\right) - 1 = (x^*)^2.
\]
There may be no $x^*$ and $y^*$ that yield this maximum (Fig.~\ref{fig:max-anc-state}), though it is a feasible solution to the general joint inference problem.
In this case, $\hat{x}_1 = x^*$ implies consistency only occurs when $x^*=0$ or $x^*=1$.
Another case,
\[
\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{3} + p_{123} + p_{23} + p_{13}\right) - 1 = \frac{1}{2} + \frac{1}{2}(x^*)^2
\]
shows $\hat{x}_1 = x^*$ implies consistency only when $x^*=1$.

@@

\begin{figure}
    \begin{minipage}{\linewidth}
        \centering
        %file: plotting/_output/invfels/ancestral_state_conditions/non-bias/plot.png
        \includegraphics[width=.95\textwidth]{analytic-anc-state}
        \label{ }
    \end{minipage}
    \begin{minipage}{\linewidth}
        \centering
        Maximal ancestral state split definitions
        \begin{tabular}{l}
        \hline
        $\hat{\boldsymbol\xi} = \{\xi_{\emptyset},\xi_{1},\xi_{2},\xi_{3},\xi_{123},\xi_{12},\xi_{23},\xi_{13}\}$\\
        \hline
        $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$\\
        $\hat{\boldsymbol\xi}_2 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\emptyset\}$\\
        $\hat{\boldsymbol\xi}_3 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\{1\}\}$\\
        $\hat{\boldsymbol\xi}_4 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\{1,2\},\{1\}\}$\\
        $\hat{\boldsymbol\xi}_5 = \{\emptyset,\emptyset,\emptyset,\{1\},\{1\},\emptyset,\{1\},\{1\}\}$\\
        \hline
        \end{tabular}
    \end{minipage}
\caption{
Regions of maximal ancestral state splits on the InvFels tree $\tau_1$.
Data generated as in Fig.~\ref{fig:bl-general-inconsistency}.
}
\label{fig:max-anc-state}
\end{figure}

@@

The regions in Fig.~\ref{fig:max-anc-state} are analytically-derived regions of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
As the region of degeneracy in Fig.~\ref{fig:max-anc-state} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
It is easy to see that when $\emptyset$ is the maximal ancestral state split, we have the same log-likelihood for $\tau_1$ and $\tau_2$.
Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the region in Fig.~\ref{fig:max-anc-state} where $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_1$.
We can see this results in the likelihood of both topologies being equal, i.e.,
\begin{align*}
\max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
\end{align*}

### 2018-10-04 joint_inf.tex
pass through comments

%EM Starting here, we only talk about branch lengths and don't mention fidelities, whereas just above we say "estimate a topology and branch fidelities using this"...
%EM At least remind the reader that we're actually talking about fidelities as the variables.
\subsection*{Inconsistent branch length estimation}
When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for almost all $x^*$ and $y^*$ in $(0,1)$ any branch length estimate is consistently biased.

@@

In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
%EM Totally confused here!
%EM 1. The generating probabilities do contain x* and y*.
%EM 2. "This is apparent" makes it sound like it should follow directly and that it doesn't require a proof.
This is apparent given Table~\ref{tab:gen-sitepatprob}, as the solution $\hat{t}$ is a linear combination of $p_{\siteSplit_j}$ values, and no generating probability contains an $x^*$ or $y^*$ term.
We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
We give an example of consistency in a degenerate case in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.

@@

%EM Is this the right figure reference? It seems strange-- it's about the max ancestral states.
This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:max-anc-state}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.

@@

\begin{figure}
\centering
%file: plotting/_output/invfels/joint_empirical/non-bias/plot.svg
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
    Data generated as in Fig.~\ref{fig:max-anc-state}.
    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
}
\label{fig:bl-general-inconsistency}
\end{figure}

>>>>

%EM Starting here, we only talk about branch lengths and don't mention fidelities, whereas just above we say "estimate a topology and branch fidelities using this"...
%EM At least remind the reader that we're actually talking about fidelities as the variables.
%das: clarification
\subsection*{Inconsistent branch parameter estimation}
When the topology is known and fixed and we estimate only the branch parameters, we show that for almost all generating parameter values, any branch parameter estimate is consistently biased.
For the branch parameter, we use branch fidelities (discussed earlier) in all statements and proofs, though these trivially extend to branch lengths via transformation.

@@

In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
%EM Totally confused here!
%EM 1. The generating probabilities do contain x* and y*.
%EM 2. "This is apparent" makes it sound like it should follow directly and that it doesn't require a proof.
%das: ok, took a stab at clarifying
For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
For the generating probabilities (top panel of Table~\ref{tab:gen-sitepatprob}), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic---$(x^*)^2$ and $(x^*)^2(y^*)^2$---or have both $x^*$ and $y^*$---$x^*(y^*)^2$.
Thus, we cannot obtain a linear combination that in the general case yields $x^*$.
We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
We give an example of consistency in a degenerate case in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.

@@

%EM Is this the right figure reference? It seems strange-- it's about the max ancestral states.
%das: ha oops! you're right
This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.

@@

\begin{figure}
\centering
%file: plotting/_output/invfels/joint_empirical/non-bias/plot.svg
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
    Regions derived in terms of probabilities of a character change along a branch for ``perfect''data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
}
\label{fig:bl-general-inconsistency}
\end{figure}
